---
title: "A quick overview of the _S4Arrays_ package"
author:
- name: Hervé Pagès
  affiliation: Fred Hutchinson Cancer Research Center, Seattle, WA
date: "Compiled `r doc_date()`;  Modified 23 March 2023"
package: S4Arrays
vignette: |
  %\VignetteIndexEntry{A quick overview of the S4Arrays package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document
---


# Introduction

`r Biocpkg("S4Arrays")` is an infrastructure package that provides a
foundation for implementing _array-like_ containers in Bioconductor,
that is, containers that mimic the behavior of ordinary matrices or
arrays in R. It is not intended to be used directly by the end user.

# The Array virtual class

At the center of this foundation is the Array virtual class whose
purpose is to be extended by other S4 classes that wish to implement
a container with an array-like semantic. Examples of such classes are:

- The SparseArray class defined in the `r Biocpkg("SparseArray")` package.

- The DelayedArray class defined in the `r Biocpkg("DelayedArray")` package.

- The ArrayGrid and ArrayViewport classes defined in the
  `r Biocpkg("S4Arrays")` package itself.

Note that Array is a virtual class with no slots:

```{r, message=FALSE}
library(S4Arrays)

showClass("Array")
```

# The extract_array() generic function

The `r Biocpkg("S4Arrays")` package also introduces the `extract_array()`
S4 generic function that Arrays subclasses (a.k.a. Arrays extensions)
are expected to support via specific methods. This allows some basic
operations like `type()`, `as.array()` or `as.matrix()` to work
out-of-the-box on instances of these Arrays subclasses (a.k.a.
Arrays derivatives). It also allows them to be used as the seed of a
DelayedArray object.

Please refer to the man page of the `extract_array()` function for more
information: `?extract_array`

# Block processing of array-like objects

The `r Biocpkg("S4Arrays")` package provides a framework that facilitates
block processing of array-like objects. This framework consists of:

- ArrayGrid and ArrayViewport objects for representing grids and viewports
  on array-like objects. See `?ArrayGrid` for more information.

- The `read_block()` and `write_block()` functions to read and write array
  blocks. See `?read_block` and `?write_block` for more information.

- The `mapToGrid()` and `mapToRef()` functions to map a set of reference
  array positions to grid positions and vice-versa. See `?mapToGrid` for
  more information.

# Other functionalities

In addition to the above, the `r Biocpkg("S4Arrays")` package provides
the following functionalities:

- The `is_sparse()` generic function for determining whether an array-like
  object uses a sparse representation or not. See `?is_sparse` for more
  information.

- Low-level utilities for manipulating array selections. See `?Lindex2Mindex`
  for more information.

- `aperm2()`: an extension of `base::aperm()` that allows dropping and/or
  adding _ineffective dimensions_. See `?aperm2` for more information.

- The `arbind()` and `acbind()` generic functions for binding array-like
  objects with an arbitrary number of dimensions along their rows or
  columns. See `?arbind` for more information.

# Session information

```{r}
sessionInfo()
```

